Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-07-30 doc/MS_v5.Rmd
overdue commits

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret A. Payseur"
output:
  word_document:
  pdf_document:
    includes:
      in_header: header.tex
  html_document:
always_allow_html: yes
bibliography: refs_Main_Report.bib
csl: elife.csl
---

@@

Sex is perhaps the most notable axis along which recombination rate varies. Broadly speaking, sexual dimorphism in the genome-wide recombination rate assumes two forms. In species such as Drosophila melanogaster, one sex completes meiosis without forming crossovers (“achiasmy”), while the other sex recombines [@huxley1928; @haldane1922sex; @burt1991]. Alternatively, in most species with recombination, crossovers occur in both sexes but at different rates (“heterochiasmy”). In these species, females tend to recombine more than males [@lorch2005; @brandvain2012scrambling; @burt1991; @lenormandDuthiel_2005; @bell1982]. In plants, heterochiasmy is correlated with the opportunity for haploid selection [@lenormandDuthiel_2005]. Despite the establishment of these interspecific trends, an understanding of how sex shapes the evolution of recombination cannot be achieved with available data.

Comprehensive comparisons of variation in female and male recombination rates within species have come from outbred populations of humans [@Kong2004; @Kong2008; @Kong2014; @halldorsson2019], dog [@wong2010dog], cattle [@ma2015_cattle; @Shen2018_cattle], and Soay sheep [@johnston2016_soay], in which the role of sex is confounded with the contributions of genetic variation. Although it is known that the level and direction of heterochiasmy can differ among species [@brandvain2012scrambling; @lenormandDuthiel_2005], the correlation between female and male recombination rates among closely related species remains poorly documented. Direct contrasts between the two sexes across a common, diverse set of genomic backgrounds that represent recent timescales would reveal whether the genome-wide recombination rate evolves differently in males and females.

@@

The house mouse, Mus musculus, provides a compelling system for understanding how sex determines the evolution of recombination. Multiple subspecies share a most recent common ancestor approximately 0.5 million years ago [@geraldes2011], providing the opportunity to examine natural variation on recent evolutionary timescales. Wild Mus musculus belong to the same species as classical inbred strains of mice, where the molecular and cellular pathways that lead to crossovers have been studied extensively [@baudat2013; @handel2010; @bolcun2012]. Single-cell immunofluorescent approaches make it possible to estimate genome-wide recombination rates in individual males and females [@koehler2002; @peters_1997]. A collection of wild-derived inbred strains founded from a variety of geographic locations is available, enabling genetic variation in recombination to be profiled across the species range. Most importantly, by measuring recombination rates in females and males from the same set of wild-derived inbred strains, the evolutionary dynamics of recombination can be directly compared in the two sexes.

@@

@@

```{r MAIN.TABLE1, echo=FALSE, eval=FALSE}

df.strain.table <- strain.table.HQ_Fig
#kable(df.strain.table)
#pander(df.strain.table)

#kable(strain.table.HQ_Fig.og, col.names = c("Subspecies", "Strain","Sex", "Number of Mice", #"Number of Cells", "Mean MLH1 count per cell", "cV", "Var", "sd", "se"), caption = "Strain #Level Summary Statistics" )  %>%
#  kable_styling() %>%
#  scroll_box(width = "800px", height = "200px")

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and three other species of Mus ( _M. spretus_, _M. spicilegus_ , and _M. caroli_ ).  Approximately 25 cells were sampled per cell were sampled given time constraints and based on previous studies [@Wang2017island; @dumont2011murid; @murdoch2010]. Mean MLH1 focus counts for `r sum(main.table$Nmice)` mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` oocytes) (Table 1).

@@

To further partition variation in recombination rate, we fit a series of linear models to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` house mice (from three subspecies; _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_) (Table 2; detailed results available in Supplemental Table X). Strain, sex, subspecies, and sex\*subspecies each affect MLH1 focus count in a linear mixed model (M1; strain (random effect): p < `r 1/10000`-- 10-4; sex: p = `r MLH1.MMlmer_results$sex.results`; subspecies: p = `r MLH1.MMlmer_results$sub.sp`; subspecies\*sex: p = `r MLH1.MMlmer_results$interaction.pval`).

The effect of subspecies is no longer significant in a model treating all factors as fixed effects (M2; musculus p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]`, molossinus p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]`), highlighting strain and sex as salient variables. Two strains exhibit strong effects on MLH1 focus count (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]`; *domesticus^LEW^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][4]`), with sex-strain interactions involving three strains (M3; *domesticus^G^* p < 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]`; *molossinus^MSM^* p < 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`; *musculus^PWD^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`). **fix the pvalue above, change the 0 to p < 10^-6**

@@

The variation in sex differences in recombination we discovered provided an opportunity to determine whether sex differences in chromatin compaction, as measured by the length of the synaptonemal complex (SC), are reversed when heterochiasmy is reversed. In all strains except _musculus^SKIVE^_, females have longer SCs than males, whether SC length was estimated as the total length across bivalents or as the length of short bivalents (t-tests; all p < 0.05, except short bivalents in _musculus^SKIVE^_, p = `r Q1_short.SC_t.test.SKIVE$p.value`).
Among short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from `r Q1SC.mus.av.short.biv_DIFF_PWD` (*musculus^PWD^*) to `r Q1SC.mus.av.short.biv_DIFF_WSB` (*domesticus^WSB^*) across strains. That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC length (M1: p = `r Q1.lmer_totSC_M1_results$sex.results`; M2: p = `r sum.Q1.glm_totSC_M2$coefficients[,4][8]`; M3: p =  `r sum.Q1.glm_totSC_M3$coefficients[,4][8]`) and short bivalent SC length (M1: p = `r Q1.lmer_short_M1_results$sex.results`; M2: p = `r summ.Q1.glm_short.SC_M2$coefficients[,4][8]`; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]`). The existence of some subspecies and strain effects on total SC length and short bivalent SC length (Supplemental Tables 5 and 8) further indicates that SC length has evolved among strains and among subspecies.

@@

@@

Males have longer normalized mean inter-focal distances (IFD~norm~) than females in seven out of eight strains (t-tests; p < 0.05), with only _musculus^KAZ^_ showing no difference (p = `r Q1.IFD.PER_t.test_KAZ$p.value`). Examination of IFD~norm~ distributions indicates that females are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values. Models treating IFD~norm~ as the dependent variable support the inference of stronger interference in males, with sex being the most significant variable (M1: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`; M2: p = `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]`; M3: p = `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]`). In contrast, there is no clear signal of sex differences in raw mean inter-focal distances (IFD~raw~) (Supplemental Table 13 and 14) across the full set of strains, whether they are considered separately or together. Visualization of normalized MLH1 foci positions on bivalents with two crossovers **(Figure 4C and Supplemental Figure X)** further suggests that interference distances vary more in females than in males, and that males display a stronger telomeric bias in the placement of the distal crossover.

@@

@@

To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of `r sum( (DMC1table.main$ncells.x) )` early zygotene and `r sum( (DMC1table.main$ncells.y) )` late zygotene spermatocytes from two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_) and three low-recombination strains (*musculus^KAZ^*, _domesticus^WSB^_, and _domesticus^G^_) (Table 3). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test; p < `r 1/1000000` -- 10-6). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = `r ttest.HighLow.Z.pval`). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision, at least in house mice.

@@

At the genetic level, the sex-specific patterns we documented indicate that some mutations responsible for the evolution of recombination rate have dissimilar phenotypic effects in the two sexes. A subset of the genetic variants associated with genome-wide recombination rate within populations of humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], Soay sheep [@johnston2016_soay], and cattle [@ma2015_cattle; @Shen2018_cattle] appear to show sex-specific properties, including opposite effects in females and males. Furthermore, inter-sexual correlations for recombination rate are weak in humans [@fledel2011] and Soay sheep [@johnston2016_soay]. Crosses between the strains we surveyed could be used to identify and characterize the genetic variants responsible for recombination rate evolution in house mice [@dumont2011murid; @Wang2017island; @wang2019_SC]. These variants could differentially affect females and males at any step in the recombination pathway. Although our DMC1 profiling was limited to males from a small number of strains (for practical reasons), our findings suggest that mutations that determine the number of double-strand breaks contribute to sex-specific evolution in the recombination rate. A study of two classical inbred strains and one wild-derived inbred strain of house mice also found a positive association between crossover number and double-strand break number in males [@baier2014].

Another implication of our results is that the connection between recombination rate and fitness differs between males and females. Little is known about whether and how natural selection shapes recombination rate in nature [@DapperPayseur2017; @Ritz2017]. @samuk2020 recently used a quantitative genetic test to conclude that an 8% difference in genome-wide recombination rate between females from two populations of Drosophila pseudoobscura was caused by natural selection. Applying similar strategies to species in which both sexes recombine, including house mice, would be a logical next step to understanding the sex-specific evolution of recombination rate.

@@

Regardless of the underlying mechanism, our results provide a rare demonstration that crossover interference can diverge over short evolutionary timescales. The notion that stronger interference can co-evolve with higher genome-wide recombination rate is supported by differences between breeds of cattle [@ma2015_cattle]. In contrast, mammalian species with stronger interference tend to exhibit lower genome-wide recombination rates [@ottoPaysuer2019; @segura2013]. The evolution of crossover interference and its relationship to changes in crossover number on the genomic scale is a topic deserving of more empirical and theoretical work.
Our findings further reveal that evolution of the genome-wide recombination rate does not require major changes in the degree of chromatin compaction. Female house mice consistently show longer SCs, even in strains with more recombination in males. Studies in mice [@lynn2002; @petkov2007] and humans [@gruhn2013; @tease2004] suggest that chromosomal axes are longer (and DNA loops are shorter) in females than males. Some authors have suggested that conserved sex differences in crossover positioning (more uniform placement in females) and interference strength (stronger interference in males) could be due to looser chromatin packing of the meiotic chromosome structure in females [@haenel2018; @petkov2007]. A cellular model designed to explain interference attributes sexual dimorphism in chromatin structure to greater cell volumes and oscillatory movements of telomeres and kinetochores in oocytes [@hulten2011_COM]. Recent work in mice connects the sparser recombination landscape in females to sex differences in the crossover maturation efficiency [@wang2017inefficient].

Our conclusions are accompanied by several caveats. First, MLH1 foci only identify interfering crossovers [@holloway2008mus81]. Although most crossovers belong to this class [@holloway2008mus81], our approach likely underestimated genome-wide recombination rates. Evolution of the number of non-interfering crossovers is a subject worth examining. A second limitation is that our investigation of crossover locations was confined to the relatively low resolution possible with immunofluorescent cytology. Positioning crossovers with higher resolution and maintaining single chromosome scale could reveal additional evolutionary patterns. Finally, the panel of inbred lines we surveyed may not be representative of recombination rate variation within and between subspecies of house mice. We considered most available wild-derived inbred lines, but house mice have a broad geographic distribution. Nevertheless, we expect our primary conclusion that recombination rate evolves in a sex-specific manner to be robust to geographic sampling because differences between females and males exist for the same set of inbred strains.

@@

We used a panel of wild-derived inbred strains of house mice (*Mus musculus*) and related murid species to profile natural genetic variation in recombination (Table 4). Since each mouse was from an inbred strain, they serve as a biological replicate. Our survey included 5 strains from *M. m. musculus*, 4 strains from *M. m. domesticus*, 2 strains from *M. m. molossinus*, 2 strains from *M. m. castaneus*, and 1 strain each from *M. spicilegus*, *M. spretus* and _M. caroli_. We subsequently denote strains by their abbreviated subspecies and name (e.g. *domesticus^WSB^*).

Mice were housed at dedicated, temperature-controlled facilities in the UW-Madison School of Medicine and Public Health, with the exception of mice from Gough Island, which were housed in a temperature-controlled facility in the UW-Madison School of Veterinary Medicine. Mice were sampled from a partially-inbred strain of Gough Island mice, approximately 6 generations of brother-sister matting. All mice were provided with ad libitum food and water. Procedures followed protocols approved by IACUC.

@@

The same dry-down spread technique was applied to both spermatocytes and oocytes, following [@peters_1997], with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used for MLH1 counts were between 5 and 12 weeks of age. Juvenile mice between 12 and 15 days of age were used for DMC1 counts. Both ovaries were collected from embryos (16-21 embryonic days) or neonates (0-48 hours after birth). Whole testes were incubated in 3ml of hypotonic solution for 45 minutes. Decapsulated ovaries were incubated in 300ul of hypotonic solution for 45 minutes. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, diluted 1:200). Slides were stored at -20\*C if not stained immediately. To visualize the structure of meiotic chromosomes, we used antibody markers for the centromere (CREST) and lateral element of the synaptonemal complex (SC) (SYCP3). Crossovers (COs) were visualized as MLH1 foci. Double strand breaks (DSBs) were visualized as DMC1 foci. The staining protocol followed [@anderson1999] and [@koehler2002]. Antibody staining and slide blocking were performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmunoResearch), 1X PBS, bovine serum albumin (Sigma), and Triton X-100 (Sigma)). Following a 30-minute blocking wash in ABD, each slide was incubated with 60ul of a primary antibody master mix for 48 hours at 37\*C. The master mix recipe contained polyclonal anti-rabbit anti-MLH1 (Calbiochem; diluted 1:50) or anti-rabbit anti-DMC1 (mix of DMC1), anti-goat polyclonal anti-SYCP3, (Abcam; diluted 1:50), and anti-human polyclonal antibody to CREST (Antibodies, Inc; diluted 1:200) suspended in ADB. Slides were washed twice in 50ml ADB before the first round of secondary antibody incubation for 12 hours at 37\*C. Alexa Fluor 488 donkey anti-rabbit IgG (Invitrgoen, location; diluted to 1:100) and Coumarin AMCA donkey anti-human IgG (Jackson ImmunoResearch; diluted to 1:200) were suspended in ADB. The last incubation of Alexa Fluor 568 donkey anti-goat (Invitrogen; diluted 1:100) was incubated at 1:100 for 2 hours at 37\* C. Slides were fixed with Prolong Gold Antifade (Invitrogen) for 24 hours after a final wash in 1x PBS. Three slides of cell spreads per mouse were prepared to serve as technical replicates for the staining protocol. Comparisons of multiple stained slides from the same mouse, showed no difference in mean MLH1 cell counts and mean cell quality.

@@

### Analysis

>>>>

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret A. Payseur"
output:
  word_document:
  pdf_document:
    keep_tex:  true
  html_document:
always_allow_html: yes
bibliography: refs_Main_Report.bib
csl: elife.csl
---

@@

Sex is perhaps the most notable axis along which recombination rate varies. Broadly speaking, sexual dimorphism in the genome-wide recombination rate assumes two forms. In species such as _Drosophila melanogaster_, one sex completes meiosis without forming crossovers (“achiasmy”), while the other sex recombines [@huxley1928; @haldane1922sex; @burt1991]. Alternatively, in most species with recombination, crossovers occur in both sexes but at different rates (“heterochiasmy”). In these species, females tend to recombine more than males [@lorch2005; @brandvain2012scrambling; @burt1991; @lenormandDuthiel_2005; @bell1982]. In plants, heterochiasmy is correlated with the opportunity for haploid selection [@lenormandDuthiel_2005]. Despite the establishment of these interspecific trends, an understanding of how sex shapes the evolution of recombination cannot be achieved with available data.

Comprehensive comparisons of variation in female and male recombination rates within species have come from outbred populations of humans [@Kong2004; @Kong2008; @Kong2014; @halldorsson2019; @gruhn2013], dog [@campbell2016], cattle [@ma2015_cattle; @Shen2018_cattle], and Soay sheep [@johnston2016_soay], in which the role of sex is confounded with the contributions of genetic variation. Although it is known that the level and direction of heterochiasmy can differ among species [@brandvain2012scrambling; @lenormandDuthiel_2005], the correlation between female and male recombination rates among closely related species remains poorly documented. Direct contrasts between the two sexes across a common, diverse set of genomic backgrounds that represent recent timescales would reveal whether the genome-wide recombination rate evolves differently in males and females.

@@

The house mouse, _Mus musculus_, is a compelling system for understanding how sex affects the evolution of recombination. Multiple subspecies share a most recent common ancestor approximately 0.5 million years ago [@geraldes2011], providing the opportunity to examine natural variation on recent evolutionary timescales. Wild _Mus musculus_ belong to the same species as classical inbred strains of mice, where the molecular and cellular pathways that lead to crossovers have been studied extensively [@baudat2013; @handel2010; @bolcun2012]. Single-cell immunofluorescent approaches make it possible to estimate genome-wide recombination rates in individual males and females [@koehler2002; @peters_1997]. A collection of wild-derived inbred strains founded from a variety of geographic locations is available, enabling genetic variation in recombination to be profiled across the species range. Most importantly, by measuring recombination rates in females and males from the same set of wild-derived inbred strains, the evolutionary dynamics of recombination can be directly compared in the two sexes.

@@

![Caption for the picture.](Figure1.jpg){ width=50% }

@@

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild-derived inbred strains sampled from three subspecies of house mice ( _M. musculus domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and three other species of Mus ( _M. spretus_, _M. spicilegus_ , and _M. caroli_ ). Mean MLH1 focus counts for `r sum(main.table$Nmice)` mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` oocytes) (Table 1).

@@

```{r MAIN.TABLE1, echo=FALSE, eval=TRUE}

#df.strain.table <- strain.table.HQ_Fig
#kable(df.strain.table)
#pander(df.strain.table)

#read excel -- then kable
Table1_MLH1 = read.xlsx2("~/MLH1repo/doc/MainTables.xlsx", header=TRUE, 1)

#might have to change DF

kable(Table1_MLH1, digits=2) %>%
  kable_styling(latex_options="scale_down") %>%
  collapse_rows(columns = 1:3)

To further partition variation in recombination rate, we fit a series of linear models to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` house mice from three subspecies; _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ (Table 2; detailed results available in Supplemental Table X). Strain, sex, subspecies, and sex\*subspecies each affect MLH1 focus count in a linear mixed model (M1; strain (random effect): p < `r 1/10000`-- 10-4; sex: p = `r MLH1.MMlmer_results$sex.results`; subspecies: p = `r MLH1.MMlmer_results$sub.sp`; subspecies\*sex: p = `r MLH1.MMlmer_results$interaction.pval`).

The effect of subspecies is no longer significant in a model treating all factors as fixed effects (M2; _musculus_ p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]`, _molossinus_ p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]`), highlighting strain and sex as salient variables. Two strains exhibit strong effects on MLH1 focus count (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]`; *domesticus^LEW^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][4]`), with sex-strain interactions involving three strains (M3; *domesticus^G^* p < 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]`; *molossinus^MSM^* p < 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`; *musculus^PWD^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`). **fix the pvalue above, change the 0 to p < 10^-6**

![Caption for the picture.](Figure2.jpg){ width=50% }

@@

The variation in sex differences in recombination we discovered provided an opportunity to determine whether sex differences in chromatin compaction, as measured by the length of the synaptonemal complex (SC), are reversed when heterochiasmy is reversed. In all strains except _musculus^SKIVE^_, females have longer SCs than males, whether SC length was estimated as the total length across bivalents or as the length of short bivalents (t-tests; all p < 0.05, except short bivalents in _musculus^SKIVE^_, p = `r Q1_short.SC_t.test.SKIVE$p.value`).
Among short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from `r Q1SC.mus.av.short.biv_DIFF_PWD` (*musculus^PWD^*) to `r Q1SC.mus.av.short.biv_DIFF_WSB` (*domesticus^WSB^*) across strains. That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC length (M1: p = `r Q1.lmer_totSC_M1_results$sex.results`; M2: p = `r sum.Q1.glm_totSC_M2$coefficients[,4][8]`; M3: p =  `r sum.Q1.glm_totSC_M3$coefficients[,4][8]`) and short bivalent SC length (M1: p = `r Q1.lmer_short_M1_results$sex.results`; M2: p < `r summ.Q1.glm_short.SC_M2$coefficients[,4][8]`; M3: p < `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]`). The existence of some subspecies and strain effects on total SC length and short bivalent SC length (Supplemental Tables 5 and 8) further indicates that SC length has evolved among strains and among subspecies.

@@

![Caption for the picture.](Fig3_SC.jpg){ width=50% }

@@

Males have longer normalized mean inter-focal distances (IFD~norm~) than females in seven out of eight strains (t-tests; p < 0.05), with only _musculus^KAZ^_ showing no difference (p = `r Q1.IFD.PER_t.test_KAZ$p.value`). Examination of IFD~norm~ distributions indicates that females are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values. Models treating IFD~norm~ as the dependent variable support the inference of stronger interference in males, with sex being the most significant variable (M1: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`; M2: p = `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]`; M3: p = `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]`). In contrast, there is no clear signal of sex differences in raw mean inter-focal distances (IFD~raw~) (Supplemental Table 13 and 14) across the full set of strains, whether they are considered separately or together. Visualization of normalized MLH1 foci positions on bivalents with two crossovers (Figure 3C and Supplemental Figure 3) further suggests that interference distances vary more in females than in males, and that males display a stronger telomeric bias in the placement of the distal crossover.

@@

```{r DMC1.Table, echo=FALSE}

#insert and read

@@

To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of `r sum( (DMC1table.main$ncells.x) )` early zygotene and `r sum( (DMC1table.main$ncells.y) )` late zygotene spermatocytes from two high-recombination strains (*musculus^PWD^* and _molossinus^MSM^_) and three low-recombination strains (*musculus^KAZ^*, _domesticus^WSB^_, and _domesticus^G^_) (Table 3). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test; p < `r 1/1000000`). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = `r ttest.HighLow.Z.pval`). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision, at least in house mice.

@@

At the genetic level, the sex-specific patterns we documented indicate that some mutations responsible for the evolution of recombination rate have dissimilar phenotypic effects in the two sexes. A subset of the genetic variants associated with genome-wide recombination rate within populations of humans [@Kong2004; @Kong2008; @Kong2014; @halldorsson2019], cattle [@ma2015_cattle; @Shen2018_cattle], and Soay sheep [@johnston2016_soay] appear to show sex-specific properties, including opposite effects in females and males. Furthermore, inter-sexual correlations for recombination rate are weak in humans [@fledel2011] and Soay sheep [@johnston2016_soay]. Crosses between the strains we surveyed could be used to identify and characterize the genetic variants responsible for recombination rate evolution in house mice [@dumont2011murid; @Wang2017island; @wang2019_SC]. These variants could differentially affect females and males at any step in the recombination pathway. Although our DMC1 profiling was limited to males from a small number of strains (for practical reasons), our findings suggest that mutations that determine the number of double-strand breaks contribute to sex-specific evolution in the recombination rate. A study of two classical inbred strains and one wild-derived inbred strain of house mice also found a positive association between crossover number and double-strand break number in males [@baier2014].

Another implication of our results is that the connection between recombination rate and fitness differs between males and females. Little is known about whether and how natural selection shapes recombination rate in nature [@DapperPayseur2017; @Ritz2017]. @samuk2020 recently used a quantitative genetic test to conclude that an 8% difference in genome-wide recombination rate between females from two populations of_Drosophila pseudoobscura_ was caused by natural selection. Applying similar strategies to species in which both sexes recombine, including house mice, would be a logical next step to understanding the sex-specific evolution of recombination rate.

@@

Regardless of the underlying mechanism, our results provide a rare demonstration that crossover interference can diverge over short evolutionary timescales. The notion that stronger interference can co-evolve with higher genome-wide recombination rate is supported by differences between breeds of cattle [@ma2015_cattle]. In contrast, mammalian species with stronger interference tend to exhibit lower genome-wide recombination rates [@ottoPaysuer2019; @segura2013]. The evolution of crossover interference and its relationship to changes in crossover number on the genomic scale is a topic deserving of more empirical and theoretical work.
Our findings further reveal that evolution of the genome-wide recombination rate does not require major changes in the degree of chromatin compaction. Female house mice consistently show longer SCs, even in strains with more recombination in males. Studies in mice [@lynn2002; @petkov2007] and humans [@gruhn2013; @tease2004] suggest that chromosomal axes are longer (and DNA loops are shorter) in females than males. Some authors have suggested that conserved sex differences in crossover positioning (more uniform placement in females) and interference strength (stronger interference in males) could be due to looser chromatin packing of the meiotic chromosome structure in females [@haenel2018; @petkov2007]. A cellular model designed to explain interference attributes sexual dimorphism in chromatin structure to greater cell volumes and oscillatory movements of telomeres and kinetochores in oocytes [@hulten2011_COM]. Recent work in mice connects the sparser recombination landscape in females to sex differences in crossover maturation efficiency [@wang2017inefficient].

Our conclusions are accompanied by several caveats. First, MLH1 foci only identify interfering crossovers [@holloway2008mus81]. Although most crossovers belong to this class [@holloway2008mus81], our approach likely underestimated genome-wide recombination rates. Evolution of the number of non-interfering crossovers is a subject worth examining. A second limitation is that our investigation of crossover locations was confined to the relatively low resolution possible with immunofluorescent cytology. Positioning crossovers with higher resolution could reveal additional evolutionary patterns. Finally, the panel of inbred lines we surveyed may not be representative of recombination rate variation within and between subspecies of house mice. We considered most available wild-derived inbred lines, but house mice have a broad geographic distribution. Nevertheless, we expect our primary conclusion that recombination rate evolves in a sex-specific manner to be robust to geographic sampling because differences between females and males exist for the same set of inbred strains.

@@

We used a panel of wild-derived inbred strains of house mice (*Mus musculus*) and related murid species to profile natural genetic variation in recombination (Table 4). Mice from the same inbred strain served as serve as a biological replicates. Our survey included 5 strains from *M. m. musculus*, 4 strains from *M. m. domesticus*, 2 strains from *M. m. molossinus*, 2 strains from *M. m. castaneus*, and 1 strain each from *M. spicilegus*, *M. spretus* and _M. caroli_. We subsequently denote strains by their abbreviated subspecies and name (*e.g. domesticus^WSB^*). Mice were housed at dedicated, temperature-controlled facilities in the UW-Madison School of Medicine and Public Health, with the exception of mice from Gough Island, which were housed in a temperature-controlled facility in the UW-Madison School of Veterinary Medicine. Mice were sampled from a partially inbred strain of Gough Island mice, approximately 6 generations of brother-sister matting. All mice were provided with *ad libitum* food and water. Procedures followed protocols approved by IACUC.

@@

The same dry-down spread technique was applied to both spermatocytes and oocytes, following @peters_1997, with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in @peterson2019. The majority of mice used for MLH1 counts were between 5 and 12 weeks of age. Juvenile males between 12 and 15 days of age were used for DMC1 counts. Both ovaries were collected from embryos (16-21 embryonic days) or neonates (0-48 hours after birth). Whole testes were incubated in 3ml of hypotonic solution for 45 minutes. Decapsulated ovaries were incubated in 300ul of hypotonic solution for 45 minutes. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, diluted 1:200). Slides were stored at -20\*C if not stained immediately. To visualize the structure of meiotic chromosomes, we used antibody markers for the centromere (CREST) and lateral element of the synaptonemal complex (SC) (SYCP3). Crossovers (COs) were visualized as MLH1 foci. Double strand breaks (DSBs) were visualized as DMC1 foci. The staining protocol followed [@anderson1999] and [@koehler2002]. Antibody staining and slide blocking were performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmunoResearch), 1X PBS, bovine serum albumin (Sigma), and Triton X-100 (Sigma)). Following a 30-minute blocking wash in ABD, each slide was incubated with 60ul of a primary antibody master mix for 48 hours at 37\*C. The master mix recipe contained polyclonal anti-rabbit anti-MLH1 (Calbiochem; diluted 1:50) or anti-rabbit anti-DMC1 (mix of DMC1), anti-goat polyclonal anti-SYCP3, (Abcam; diluted 1:50), and anti-human polyclonal antibody to CREST (Antibodies, Inc; diluted 1:200) suspended in ADB. Slides were washed twice in 50ml ADB before the first round of secondary antibody incubation for 12 hours at 37\*C. Alexa Fluor 488 donkey anti-rabbit IgG (Invitrgoen, location; diluted to 1:100) and Coumarin AMCA donkey anti-human IgG (Jackson ImmunoResearch; diluted to 1:200) were suspended in ADB. The last incubation of Alexa Fluor 568 donkey anti-goat (Invitrogen; diluted 1:100) was incubated at 1:100 for 2 hours at 37\* C. Slides were fixed with Prolong Gold Antifade (Invitrogen) for 24 hours after a final wash in 1x PBS. Three slides of cell spreads per mouse were prepared to serve as technical replicates for the staining protocol. Comparisons of multiple stained slides from the same mouse showed no difference in mean MLH1 cell counts and mean cell quality. Sampled numbers of mice and cells per mouse were maximized to the extent possible given constraints on breeding and time.

@@

### Analyses

To estimate the number of crossovers across the genome, we counted MLH1 foci within bivalents, synapsed homologous chromosomes. MLH1 foci were counted in pachytene cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score ranging from 1 (best) to 5 (worst) was assigned to each cell based on visual appearance of staining and spread of bivalents. Cells with a score of 5 were excluded from the final analysis. Distributions of MLH1 count per cell were visually inspected for normality (Supplemental Figure 1). When outliers for MLH1 count were found during preliminary analysis, the original images were inspected and the counts confirmed.
